---
title: "Covariates"
---

# Overview 

We are generating predictions at the grid cell scale (10x10km) using a set of covariates from the latest NABat report. We are focusing on the state of Montana, and incorporating different jurisdictions that may or may not implement all management actions. 

## Data sources 
Covariate data and grids are coming from NABat 

```{r libs}
#| message: false

# Libraries
library(tidyverse)
library(sf)
library(jagsUI)
library(scales)

```

```{r load_data}

## COVARIATES --------
mt_covariates <- readRDS("datafiles/mt_covariates.RDS") %>% 
  # Abbreviate jurisdiction names
    mutate(jurisd_abbr = case_when(
    jurisdiction == "Montana Fish, Wildlife, and Parks" ~ "MFWP",
    jurisdiction == "US Bureau of Land Management" ~ "BLM",
    jurisdiction == "US Forest Service" ~ "USFS",
    jurisdiction == "Reservations" ~ "Res",
    jurisdiction == "US Fish and Wildlife Service" ~ "USFWS",
    jurisdiction == "National Park Service" ~ "NPS",
    jurisdiction == "Montana Department of Natural Resources and Conservation" ~ "DNRC"
    ))

grid_geometry <- mt_covariates %>% 
  select(GRTS_ID)
```

```{css, echo=FALSE}
.panel-tabset .nav-item {
  font-size: 12px;
}
```


# Covariates 
All the information from the report regarding the different covariates they used for their analysis.

## Ecoregions

:::{.panel-tabset}

## Eco1
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = eco1_name))
```

## Eco2
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = eco2_name))
```

## Eco3
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = eco3_name))
```
:::

## Covariates

:::{.panel-tabset}
## Elev
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = DEM_max))
```

## Temp
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = mean_temp))
```

## Physdiv
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = physio_div))
```

## Precip
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = precip))
```

## Forest
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = p_forest))
```

## Wetland
```{r}
ggplot(data = mt_covariates) + geom_sf(aes(fill = p_wetland))
```


:::
