---
title: "Estimation"
execute: 
  echo: false
html:
    code-fold: true
    code-summary: "Show the code"
---

```{r libraries}
#| message: false
library(tidyverse)
library(jagsUI)
library(sf)
```

```{r load_calls}
#| messgae: false
#| warning: false
raw_wns_calls <- read_csv("datafiles/mt_batcalls/mt_calls_csvs/WNSCallsCapHistory.csv", show_col_types = FALSE)

# Transform to long format:
raw_wns_calls %>%
  # create variable for julian date, formerly in the columns
  pivot_longer(., 4:last_col(), names_to = "jdate", values_to = "y") %>%
  # keep only needed variables
  select(SiteID, CSY, jdate, y) %>%
  # julian date should be a number
  mutate(jdate = as.double(jdate)) %>%
  # remove NAs - means recorders wasn't on
  drop_na(y) %>%
  # clean the names
  janitor::clean_names() %>%
  # separate CSY into three
  separate(csy, c("cell", "site", "year"), remove = FALSE) %>%
  # create index for siteid.year, and cell.site
  mutate(site_id_year = paste(site_id, year, sep = "_"),
         cell_site = paste(cell, site, sep = "_"),
         cell_year = paste(cell, year, sep = "_")) %>%
  # create binary version of detection data
  mutate(det_non_det = as.numeric(y > 0)) %>%
  # Create state categories, y multi state
  # Asumming a cutoff of 100 bat calls as many bats
  mutate(y_multi_state = case_when(
    y == 0 ~ 1,
    y <= 100 ~ 2,
    y > 100 ~ 3
  )) %>%
  # add column for number of nights of recording for each site_id
  add_count(site_id, name = "n_nights_site_id") -> wns_calls_long

```


```{r site_covariates}
#| message: false
#| warning: false
#| results: hide
raw_sites <- read_csv("datafiles/mt_batcalls/mt_calls_csvs/SiteDataCleanFinal.csv")
```


```{r mt_covariates}
mt_covariates <- readRDS("datafiles/mt_covariates.RDS")
```

# Multiscale occupancy

## Model description

General model based on occupancy at the grid cell level, each with several sites within each cell, and multiple surveys to each site:

$$
z_i \sim \text{Bernoulli}(\psi_i)
$$

$$
u_{i,j} \sim \text{Bernoulli}(z_i \times \theta_{i,j})
$$

$$
y_{i,j,k} \sim \text{Bernoulli}(u_{i,j} \times p_{i,j,k})
$$
We incorporate covariates as follows, with the probability of occupancy ($\psi_i$) for each site $i$ given by:

$$
\begin{align*}
logit(\psi_i) = \beta_0 + \beta_1 \cdot \text{Elevation} + \beta_2 \cdot \text{Elevation}^2 +
\beta_3 \cdot \text{Temp} + \beta_4 \cdot \text{Temp}^2 + \\
\beta_5 \cdot \text{Physiographic diversity} +
\beta_6 \cdot \text{Precipitation} + \\
\beta_7 \cdot \text{Percent Forest} + \beta_8 \cdot \text{Percent Wetlands}
\end{align*}
$$

local occupancy ($u_{i,j}$) at each site $j$ within grid cell $i$ dependent on:

$$
logit(\theta_{i,j}) = \alpha_0 + \alpha_1 \cdot \text{Start date} + \alpha_2 \cdot \text{Duration} + \alpha_3 \cdot \text{Site type}
$$
I did not include covariates for the detection probability.

## Code

### Bunde data

```{r}

```




# Multistate mean occupancy

# Dynamic multistate mean occupancy

# Multistate multiscale occupancy

# Dynamic multistate multiscale occupancy